---
title: "Macro Analysis of Immigration and the Labor Force"
author: "Jared White"
date: "`r Sys.Date()`"
output: github_document
---
##### License: [CC-BY-4.0](https://creativecommons.org/licenses/by/4.0/deed.en)

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(fredr)
library(lubridate)
library(scales)
# Check for key in environment. FRED API key must be present.
ifelse(
  fredr_has_key() == TRUE,
  paste0("FRED API key is available."),
  paste0("FRED API key is NOT available. A valid FRED API Key is required to download the needed datasets.")
)

```

```{r message=FALSE, warning=FALSE, include=FALSE}
# fetch employment rates
f_empl <- fredr(
  series_id = "LNU02073395",
  units = "pc1",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)

n_empl <- fredr(
  series_id = "LNU02073413",
  units = "pc1",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# fetch participation rates
f_part <- fredr(
  series_id = "LNU01373395",
  units = "lin",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)

n_part <- fredr(
  series_id = "LNU01373413",
  units = "lin",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# fetch population levels
f_pop <- fredr(
  series_id = "LNU00073395",
  units = "pc1",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)

n_pop <- fredr(
  series_id = "LNU00073413",
  units = "pc1",
  frequency = "q",
  observation_start = as.Date("2008-01-01"),
  observation_end = as.Date("2024-03-01")
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# fetch wages by education level and population group
f_wage_lhs <- fredr(
  series_id = "LEU0257370100A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
n_wage_lhs <- fredr(
  series_id = "LEU0257375500A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
f_wage_lc <- fredr(
  series_id = "LEU0257370200A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
n_wage_lc <- fredr(
  series_id = "LEU0257375600A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
f_wage_c <- fredr(
  series_id = "LEU0257370400A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
n_wage_c <- fredr(
  series_id = "LEU0257375800A",
  units = "lin",
  frequency = "a",
  observation_start = as.Date("2019-01-01"),
  observation_end = as.Date("2023-01-01")
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# calculate pct change by group from 2020 for each table
f_lhs <- 
  c("foreign-born", "No High School Diploma",
    round(
      as.numeric(
        (f_wage_lhs$value[f_wage_lhs$date == "2023-01-01"] /
        f_wage_lhs$value[f_wage_lhs$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(f_lhs) <- c("group","education","percent_change")

n_lhs <- 
  c("native-born", "No High School Diploma",
    round(
      as.numeric(
        (n_wage_lhs$value[n_wage_lhs$date == "2023-01-01"] /
        n_wage_lhs$value[n_wage_lhs$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(n_lhs) <- c("group","education","percent_change")

f_lc <- 
  c("foreign-born", "No College Degree",
    round(
      as.numeric(
        (f_wage_lc$value[f_wage_lc$date == "2023-01-01"] /
        f_wage_lc$value[f_wage_lc$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(f_lc) <- c("group","education","percent_change")

n_lc <- 
  c("native-born", "No College Degree",
    round(
      as.numeric(
        (n_wage_lc$value[n_wage_lc$date == "2023-01-01"] /
        n_wage_lc$value[n_wage_lc$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(n_lc) <- c("group","education","percent_change")

f_c <- 
  c("foreign-born", "Bachelors or Higher",
    round(
      as.numeric(
        (f_wage_c$value[f_wage_c$date == "2023-01-01"] /
        f_wage_c$value[f_wage_c$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(f_c) <- c("group","education","percent_change")

n_c <- 
  c("native-born", "Bachelors or Higher",
    round(
      as.numeric(
        (n_wage_c$value[n_wage_c$date == "2023-01-01"] /
        n_wage_c$value[n_wage_c$date == "2020-01-01"]) * 100 - 100
      ),digits = 2)
  )
names(n_c) <- c("group","education","percent_change")
#combine tables and cast proper datatype.
wage_change <- bind_rows(f_lhs, n_lhs, f_lc, n_lc, f_c, n_c)
wage_change$percent_change <- as.numeric(wage_change$percent_change)
#table for 2023 wages
wage_d <- data.frame(
  group = c("foreign-born","native-born","difference"),
  lhs = c(f_wage_lhs$value[f_wage_lhs$date == "2023-01-01"],
          n_wage_lhs$value[n_wage_lhs$date == "2023-01-01"],
            f_wage_lhs$value[f_wage_lhs$date == "2023-01-01"]-
            n_wage_lhs$value[n_wage_lhs$date == "2023-01-01"]),
  lc = c(f_wage_lc$value[f_wage_lc$date == "2023-01-01"],
         n_wage_lc$value[n_wage_lc$date == "2023-01-01"],
            f_wage_lc$value[f_wage_lc$date == "2023-01-01"]-
            n_wage_lc$value[n_wage_lc$date == "2023-01-01"]),
  c = c(f_wage_c$value[f_wage_c$date == "2023-01-01"],
        n_wage_c$value[n_wage_c$date == "2023-01-01"],
            f_wage_c$value[f_wage_c$date == "2023-01-01"]-
            n_wage_c$value[n_wage_c$date == "2023-01-01"])
)
#clean up unneeded dataframes
remove(f_wage_lhs,n_wage_lhs,f_wage_lc,n_wage_lc,f_wage_c,n_wage_c,
       f_lhs,n_lhs,f_lc,n_lc,f_c,n_c)
```

### Bottom Line Upfront

- **Growth in Employment** is more volatile and strongly correlated with overall U.S. economic health in the foreign-born population category than native-born.

- **Labor Force Participation Rates** are consistently higher in the foreign-born population category and recover faster from disincentive to employment.

- **Growth in Total Population** in the foreign-born category outpaces that of the native-born. Population Growth shows correlation with employment growth in the foreign-born category, but not the native-born. Labor Force Participation shows no correlation to employment in either category.

- **Wages and Salaries** are converging between the two population categories across all education levels, with the disparity in wages smallest among college-educated individuals, indicating faster wage equalization for native-born workers.

### Employment Levels

##### Foreign Born is defined as:

*"The foreign born are persons who reside in the United States but who were not U.S. citizens at birth. Specifically, they were born outside the United States (or one of its outlying areas such as Puerto Rico or Guam), and neither parent was a U.S. citizen. The foreign born include legally-admitted immigrants, refugees, temporary residents such as students and temporary workers, and undocumented immigrants. However, the survey does not separately identify persons in these categories. For further information about the survey, see the Technical Note in this news release."* [(Bureau of Labor Statistics)](https://www.bls.gov/news.release/forbrn.htm)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# make a palette for consistent coloring.
pal_fn_empl <- c('purple','green')
names(pal_fn_empl) <- c("foreign-born","native-born")

# plot employment growth
ggplot() + 
  geom_line(data = f_empl, mapping = aes(x=date,y=value,color ="foreign-born"),
            show.legend = TRUE) +
  geom_line(data = n_empl, mapping = aes(x=date,y=value,color = "native-born"),
            show.legend = TRUE) +
  scale_color_manual(values = pal_fn_empl) +
  scale_x_date(breaks = date_breaks("1 year"),
               labels = date_format("'%y")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  labs(
    color = "Population Group",
    title = "Growth in Recorded Number of Employed People",
    subtitle = "(Percent Change from Prior Year, Quarterly Average)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 1 citations)",
    x = "Year",
    y = "Number of Employees (Yearly Change, Percent)"
  )
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# 6 Number Statistics (Quartiles and SD)
f_empl_6num <- fivenum(f_empl$value)
f_empl_sd <- sd(f_empl$value)
f_empl_6num$sd <- f_empl_sd
names(f_empl_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")
n_empl_6num <- fivenum(n_empl$value)
n_empl_sd <- sd(n_empl$value)
n_empl_6num$sd <- n_empl_sd
names(n_empl_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")

# add useful column to make color coordination and labeling easier
f_empl$group <- as.character("foreign-born")
n_empl$group <- as.character("native-born")
#boxplot
ggplot()+
  geom_boxplot(data = f_empl,aes(x = value,y = group,fill = group),
               show.legend = FALSE)+
  geom_boxplot(data = n_empl,aes(x = value, y = group,fill = group),
               show.legend = FALSE)+
  scale_fill_manual(values = pal_fn_empl)+
  scale_x_continuous(breaks = seq(-20,15, by = 5))+
  labs(
    title = "Employment Change Distribution",
    subtitle = "(Quartiles and Outliers of Percent Changes from Prior Year)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 1 citations)",
    x = "Percent Change Values in Employment",
    y = "Population Group"
  )+
  theme(axis.text.y = element_text(angle = 90,hjust = .55))
```

##### Foreign-Born Employment Change Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(as_data_frame(f_empl_6num), format = "html", 
             align = 'c',digits = 2)
```

##### Native-Born Employment Change Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(as_data_frame(n_empl_6num), format = "html", 
             align = 'c',digits = 2)
```

##### Analyst Comment:

It's important to note that these statistics do not differentiate between subgroups of foreign-born individuals in the U.S. The percentages indicate the degree of growth or decline in the total number of recorded employed individuals from the same quarter of the previous year.

This chart illustrates that while the growth in the number of employed foreign-born individuals typically outpaces that of native-born individuals during periods of economic stability, foreign-born employment declines more rapidly and profoundly during recessions (e.g., in 2008 and 2020). Conversely, foreign-born employment also increases more quickly during economic recovery from recession. In other words, foreign-born employment exhibits greater volatility and a stronger correlation with economic health compared to native-born employment.

It's essential to recognize that correlation does not imply causation. Based on the available data, there are several equally plausible insights that are **not mutually exclusive**:

  1. The overall health and outlook of the broader U.S. economy have a more significant impact on the employment of foreign-born individuals than on native-born individuals.
  
  2. Employment of foreign-born individuals serves simultaneously as both a pillar and an indicator of overall U.S. economic health.
  
  3. Since the category of foreign-born workers encompasses immigrants of all legal and employable statuses (work visa, sponsorship, naturalized, documented/undocumented, etc.), the data also suggests that this population group may be perceived by many employers as simultaneously more expendable and more readily attainable than native-born employees.
  
  4. Foreign-born employees may also be disproportionately predisposed to work in sectors that are more profoundly affected by the overall health of the economy.
  
### Labor Force Participation Rate

##### Labor Force Participation is:

*"The Bureau of Labor Statistics (BLS), a federal agency that measures labor market activity, working conditions, price changes and productivity, defines the labor force participation (LFP) rate as the number of people who are either working or actively seeking work as a share of the working age population. The working age population consists of people 16 years and over."* [(Federal Reserve Bank of St. Louis)](https://www.stlouisfed.org/open-vault/2020/august/labor-force-participation-rate-explained)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot LFP
ggplot()+
  geom_line(data = f_part, mapping = aes(x=date,y=value,color="foreign-born"),
            show.legend = TRUE)+
  geom_line(data = n_part, mapping = aes(x=date,y=value,color="native-born"),
            show.legend = TRUE)+
  scale_color_manual(values = pal_fn_empl)+
   scale_x_date(breaks = date_breaks("1 year"),
               labels = date_format("'%y")) +
  scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  labs(
    color = "Population Group",
    title = "Labor Force Participation Rate",
    subtitle = "(Percentage of Population Group, Quarterly Average)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 2 citations)",
    x = "Year",
    y = "Participation Rate (Percent)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
f_part_6num <- fivenum(f_part$value)
f_part_sd <- sd(f_part$value)
f_part_6num$sd <- f_part_sd
names(f_part_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")
n_part_6num <- fivenum(n_part$value)
n_part_sd <- sd(n_part$value)
n_part_6num$sd <- n_part_sd
names(n_part_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")

f_part$group <- "foreign-born"
n_part$group <- "native-born"

ggplot()+
  geom_boxplot(data = f_part, aes(x = value, y = group, fill = group),
               show.legend = FALSE)+
  geom_boxplot(data = n_part, aes(x = value, y= group, fill = group),
               show.legend = FALSE)+
  scale_fill_manual(values = pal_fn_empl)+
  scale_x_continuous(breaks = seq(60,70, by = 1))+
  labs(
    title = "Labor Force Participation Distribution",
    subtitle = "(Quartiles and Outliers of LFP Rates)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 2 citations)",
    x = "Values of LFP",
    y = "Population Group"
  )+
  theme(axis.text.y = element_text(angle = 90,hjust = .55))
```

##### Foreign-Born LFP Rate Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(as_data_frame(f_part_6num), format = "html",
             align = 'c',digits = 2)
```

##### Native-Born LFP Rate Numeric Summary:

```{r echo=FALSE, message=TRUE, warning=FALSE}
knitr::kable(as_data_frame(n_part_6num), format = "html"
             , align = 'c',digits = 2)
```

##### Analyst Comment:

While both population groups exhibit similar deviations and ranges in their Labor Force Participation Rates(LFP), the foreign-born category displays a consistently higher LFP compared to the native-born population. Notably, the outlying minimum of foreign-born LFP is 0.3% higher than the median of the native-born group. In striking inverse similarity, the maximum LFP for the native-born group is 0.3% lower than the foreign-born median.

The data from 2020 onward suggests that foreign-born workers are much quicker to either return to work or seek reemployment following a mass disincentive to employment. By the second quarter of 2022, the LFP of the foreign-born population had recovered to pre-pandemic levels, whereas native-born LFP has only partially reversed its total pandemic-induced decrease to date.

### Total Population Growth

##### Measurement of Population:

*"Civilian noninstitutional population is defined as persons 16 years of age and older residing in the 50 states and the District of Columbia, who are not inmates of institutions (e.g., penal and mental facilities, homes for the aged), and who are not on active duty in the Armed Forces."* [(Federal Reserve Bank of St. Louis)](https://fred.stlouisfed.org/series/LNU00073413#0)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# plot population growth
ggplot()+
  geom_line(data = f_pop, mapping = aes(x=date,y=value,color="foreign-born"),
            show.legend = TRUE)+
  geom_line(data = n_pop, mapping=aes(x=date,y=value,color="native-born"),
            show.legend = TRUE)+
  scale_color_manual(values = pal_fn_empl)+
   scale_x_date(breaks = date_breaks("1 year"),
               labels = date_format("'%y")) +
  labs(
    title = "Growth in Recorded Total Population Groups",
    subtitle = "(Percent Change from Prior Year, Quarterly Average)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 3 citations)",
    color = "Population Group",
    x = "Year",
    y = "Size of Population (Yearly Change, Percent)")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
#make 6 num stats
f_pop_6num <- fivenum(f_pop$value)
f_pop_sd <- sd(f_pop$value)
f_pop_6num$sd <- f_pop_sd
names(f_pop_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")
n_pop_6num <- fivenum(n_pop$value)
n_pop_sd <- sd(n_pop$value)
n_pop_6num$sd <- n_pop_sd
names(n_pop_6num) <- c("Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Standard Deviation")

f_pop$group <- "foreign-born"
n_pop$group <- "native-born"
#make box plots
ggplot()+
  geom_boxplot(data = f_pop, aes(x=value,y=group,fill=group),
               show.legend = FALSE)+
  geom_boxplot(data = n_pop,aes(x=value,y=group,fill=group),
               show.legend = FALSE)+
  scale_fill_manual(values = pal_fn_empl)+
  scale_x_continuous(breaks = seq(-3,6, by = 1))+
  labs(
    title = "Population Growth Distribution",
    subtitle = "(Quartiles and Outliers of Percent Changes from Prior Year)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 3 citations)",
    x = "Percent Change Values in Total Population",
    y = "Population Group"
  )+
  theme(axis.text.y = element_text(angle = 90,hjust = .55))
```

##### Foreign-Born Population Growth Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
#print 6 number stats
knitr::kable(as_data_frame(f_pop_6num),format = "html",
             align ="c",digits = 2)
```

##### Native-Born Population Growth Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}
knitr::kable(as_data_frame(n_pop_6num),format = "html",
             align = "c",digits = 2)
```

##### Analyst Comment:

Given the intrinsic link between immigration and public policy, it's not surprising that total population growth is much more volatile in the foreign-born population category. However, what is notable is that total population growth in the foreign-born category is the only metric correlated with employment growth in the same category.

Comparing chart 3 with chart 1 reveals that both employment and total population decrease in the foreign-born category during economic recession. In contrast, the growth of the total native-born population shows a slight increase during periods of decreased employment.

From the available data, it's equally plausible that:
  
  1. Individuals in the foreign-born population group are, on average, more likely to leave the measured population when economic opportunities decrease compared to their native-born counterparts.
  
  2. A decrease in the availability of foreign-born employees and workers exacerbates or causes underlying issues of economic instability.

##### Correlations of Population Growth and LFP with Employment Growth:

```{r echo=FALSE, message=FALSE, warning=FALSE}
corrs <- data.frame(group = c("foreign-born","native-born"),
                   population_growth = c(
                     cor(x = f_pop$value, y = f_empl$value),
                     cor(x = n_pop$value, y = n_empl$value)
                                              ),
                   lfp = c(
                     cor(x = f_part$value, y = f_empl$value),
                     cor(x = n_part$value, y = n_empl$value)
                                              )
                  )
knitr::kable(corrs, format = "html", digits = 2, align = "c")
```

Foreign-born total population growth is the only metric that exhibits a significant correlation score with employment growth (correlation of 0.72). Native-born population growth shows a very slight negative correlation with employment growth. Labor Force Participation Rates show no correlation with employment growth in either population group.

### Differences in Metrics between Population Groups Over Time

$$
\large Plot=\frac{Foreign-Native}{Time}
$$

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Difference in LFP between groups
lfp_diff <- data.frame(f_empl$date, (f_part$value - n_part$value))
lfp_diff$metric <- "LFP"
names(lfp_diff) <- c("date", "diff", "metric")
# Difference in Pop Groups
pop_diff <- data.frame(f_pop$date, f_pop$value - n_pop$value)
pop_diff$metric <- "Population Growth"
names(pop_diff) <- c("date", "diff", "metric")
# Difference in Employment
empl_diff <- data.frame(f_empl$date, f_empl$value - n_empl$value)
empl_diff$metric <- "Employment Growth"
names(empl_diff) <- c("date", "diff", "metric")
#palette
pal_diff <- c("red","blue","gold")
names(pal_diff) <- c("LFP","Population Growth","Employment Growth")
#plot
ggplot()+
  geom_smooth(data = lfp_diff, aes(x = date,y = diff,
                                   color="LFP"),
              method = "gam",
              show.legend = TRUE)+
  geom_point(data = lfp_diff, aes(x = date, y=diff),
             color = "#ffa6a2")+
  geom_smooth(data = pop_diff, aes(x = date,y = diff,
                                   color="Population Growth"),
              method = "gam",
              show.legend = TRUE)+
  geom_point(data = pop_diff, aes(x = date, y=diff),
             color = "#9ecae1")+
  geom_smooth(data = empl_diff, aes(x = date, y = diff,
                                    color = "Employment Growth"),
              method = "gam",
              show.legend = TRUE)+
  geom_point(data = empl_diff,aes(x = date, y=diff),
             color = "#efff85")+
  scale_color_manual(values = pal_diff)+
  scale_x_date(breaks = date_breaks("1 year"),
               labels = date_format("'%y")) +
  labs(color = "Metric Differences",
       title = "Amount Foreign LFP, Pop. Growth, and Emp. Growth Exceeds Native",
       subtitle = "(Percentages, Foreign-Born Minus Native-Born, Gam Smoothing)",
       caption = "Source: U.S. Bureau of Labor Statistics (chart 1, 2 and 3 citations)",
       x = "Year",
       y = "Difference Between Population Groups")
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot()+
  geom_boxplot(data = empl_diff,aes(x = diff,y=metric,fill=metric),
               show.legend = FALSE)+
  geom_boxplot(data = lfp_diff,aes(x=diff,y=metric,fill=metric),
               show.legend = FALSE)+
  geom_boxplot(data = pop_diff,aes(x=diff,y=metric,fill=metric),
               show.legend = FALSE)+
  scale_fill_manual(values = pal_diff)+
  scale_x_continuous(breaks = seq(-6,6, by = 1))+
  labs(
    title = "Metric Differences Distribution",
    subtitle = "(Quartiles and Outliers of Differences Between Population Groups)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart1, 2, and 3 citations)",
    x = "Difference Values in Percentages",
    y = "Metric"
  )+
  theme(axis.text.y = element_text(angle = 90,hjust = .55))
```

##### Metric Differences Numeric Summary:

```{r echo=FALSE, message=FALSE, warning=FALSE}

#metric_diff   Min   Q1    Median    Q3    Max   Empl-Growth-Corr.
#lfp
#Pop-Gr

lfp_diff_6num <- fivenum(lfp_diff$diff)
lfp_diff_6num$corr <- round(
  cor(x = lfp_diff$diff, y = empl_diff$diff),
  digits = 2)
lfp_diff_6num$metric <- "LFP"
names(lfp_diff_6num) <- c("Min","Q1",
                        "Med","Q3",
                        "Max","Corr","metric")

pop_diff_6num <- fivenum(pop_diff$diff)
pop_diff_6num$corr <- round(
  cor(x = pop_diff$diff, y = empl_diff$diff),
  digits = 2)
pop_diff_6num$metric <- "Pop. Growth"
names(pop_diff_6num) <- c("Min","Q1",
                        "Med","Q3",
                        "Max","Corr","metric")

diff_6num <- data.frame(metric=c(lfp_diff_6num$metric,
                                  pop_diff_6num$metric),
                        min=c(lfp_diff_6num$Min,
                                 pop_diff_6num$Min),
                        q1=c(lfp_diff_6num$Q1,
                             pop_diff_6num$Q1),
                        med=c(lfp_diff_6num$Med,
                              pop_diff_6num$Med),
                        q3=c(lfp_diff_6num$Q3,
                              pop_diff_6num$Q3),
                        max=c(lfp_diff_6num$Max,
                              pop_diff_6num$Max),
                        corr=c(lfp_diff_6num$Corr,
                               pop_diff_6num$Corr)
                        )
names(diff_6num) <- c("Metric","Minimum -","1st Quartile -",
                        "Median -","3rd Quartile -",
                        "Maximum -","Corr with Diff in Empl Growth")

knitr::kable(diff_6num, digits = 2, format = "html", align = "c")
```

##### Analyst Comment:

This plot displays the differences in Employment Growth, Total Population Growth, and LFP between the two population categories. It demonstrates the amount by which the foreign-born category exceeded the native-born category in each metric. Positive points indicate that the foreign-born category experienced either a higher growth rate or a slower decline rate for that metric and time period. Conversely, values below zero indicate that either the growth of the native-born category exceeded that of the foreign-born, or that a decrease occurred at a lesser rate for the native-born. 

The difference between the two population categories shows a stronger correlation of both Total Population Growth and LFP with Growth in Employment. However, it is evident that the change in total population is more strongly correlated with change in employment than the Labor Force Participation Rate. 

### Wage Growth

##### Measurement of Weekly Earnings:

*"Data measure usual weekly earnings of wage and salary workers. Wage and salary workers are workers who receive wages, salaries, commissions, tips, payment in kind, or piece rates. The group includes employees in both the private and public sectors but, for the purposes of the earnings series, it excludes all self-employed persons, both those with incorporated businesses and those with unincorporated businesses."* [(Federal Reserve Bank of St. Louis)](https://fred.stlouisfed.org/series/LEU0257370400A)

```{r echo=FALSE, message=FALSE, warning=FALSE}
# wage growth chart
ggplot(data = wage_change)+
  geom_col(aes(x=group, y=percent_change, fill = group))+
  scale_fill_manual(values = pal_fn_empl)+
  facet_wrap(~factor(education, levels = c('No High School Diploma',
                                           'No College Degree',
                                           'Bachelors or Higher')))+
  labs(
    title = "Wage and Salary Growth 2020 - 2023",
    subtitle = "(Percent Increase in Median Weekly Earnings Since 2020)",
    caption = "Source: U.S. Bureau of Labor Statistics (chart 4 citations)",
    x = "Population Group, Education Level",
    y = "Weekly Earnings Increase (Percent)",
    fill = "Population Group"
  )
```

##### 2023 Median Weekly Earnings (Dollar Amounts):

```{r echo=FALSE, message=FALSE, warning=FALSE}
names(wage_d) <- c("Population Group -","No High School Diploma -",
                   "No College Degree -","Bachelors or Higher -")
knitr::kable(wage_d,format = "html",align = "c")
```


##### Analyst Comments:

In education categories where the median wage is higher for native-born workers, foreign-born wages have shown a slightly higher percentage increase since 2020 compared to native-born workers. College-educated native-born workers' weekly earnings demonstrate a significantly higher percentage increase than their foreign-born educated counterparts. Interestingly, this is the only education category in which foreign-born workers' median earnings surpass those of native-born workers in dollar amount. 

Additionally, the disparity in wages among college-educated individuals is the smallest among all education levels. This means that, while wages are converging between the two population categories across all levels of education, they are doing so at a much faster rate for native-born workers.

### Conclusion

Economics is one aspect of the broader national discourse on immigration. Further complicating this issue is the lack of transparency and granularity of the statistics provided by the Bureau of Labor Statistics. It is impossible to discern what proportion of the recorded foreign-born workers are legally working or residing in the U.S. Since this category also encompasses undocumented residents, the extent to which the BLS sample represents the entire population group remains uncertain.

What can be concluded, however, is that with a consistently higher Labor Force Participation Rate and population growth that correlates more strongly with employment levels, the foreign-born population category serves as an essential factor in the labor market and broader economy. While, under the right circumstances, any population group can theoretically exert downward pressure on wages, the available data does not support the theory that this applies universally to foreign-born workers.

If the current trends continue, the U.S. economy will likely experience:

  - A continued decline in native-born Labor Force Participation relative to that of the foreign-born category.
  
  - A greater proportion of foreign-born individuals within the total population.
  
  - An Increasing disparity in employment growth between the two population categories.
  
  - Rising wages and salaries that also become more equitable across all population segments.

### Citations

#### Citations for Employment Level (chart 1):

U.S. Bureau of Labor Statistics, Employment Level - Foreign Born [LNU02073395], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU02073395, April 5, 2024. 

U.S. Bureau of Labor Statistics, Employment Level - Native Born [LNU02073413], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU02073413, April 5, 2024.

#### Citations for Labor Force Participation Rates (chart 2):

U.S. Bureau of Labor Statistics, Labor Force Participation Rate - Foreign Born [LNU01373395], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU01373395, April 6, 2024. 

U.S. Bureau of Labor Statistics, Labor Force Participation Rate - Native Born [LNU01373413], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU01373413, April 6, 2024. 

#### Citations for Total Population (chart 3):

U.S. Bureau of Labor Statistics, Population Level - Foreign Born [LNU00073395], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU00073395, April 7, 2024. 

U.S. Bureau of Labor Statistics, Population Level - Native Born [LNU00073413], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LNU00073413, April 7, 2024.

#### Citations for Wage Growth (chart 4):

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: Less than a High School diploma: 25 years and over: Foreign born [LEU0257370100A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257370100A, April 7, 2024. 

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: Less than a High School diploma: 25 years and over: Native born [LEU0257375500A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257375500A, April 7, 2024. 

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: High School graduates, No Degree: 25 years and over: Foreign born [LEU0257370200A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257370200A, April 7, 2024. 

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: High School graduates, No Degree: 25 years and over: Native born [LEU0257375600A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257375600A, April 7, 2024. 

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: Bachelor's degree and higher: 25 years and over: Foreign born [LEU0257370400A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257370400A, April 7, 2024. 

U.S. Bureau of Labor Statistics, Employed full time: Median usual weekly nominal earnings (second quartile): Wage and salary workers: Bachelor's degree and higher: 25 years and over: Native born [LEU0257375800A], retrieved from FRED, Federal Reserve Bank of St. Louis; https://fred.stlouisfed.org/series/LEU0257375800A, April 7, 2024.

#### Citations for R and Libraries

R Core Team (2021). R: A language and environment for statistical computing.
  R Foundation for Statistical Computing, Vienna, Austria. 
  URL https://www.R-project.org/.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R, Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E, Bache SM, Müller K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi K, Vaughan D, Wilke C, Woo K, Yutani H (2019).
  “Welcome to the tidyverse.” _Journal of Open Source Software_, *4*(43), 1686.
  doi: 10.21105/joss.01686 (URL: https://doi.org/10.21105/joss.01686).

Sam Boysel and Davis Vaughan (2021). fredr: An R Client for the 'FRED' API. R package
  version 2.1.0. https://CRAN.R-project.org/package=fredr

Garrett Grolemund, Hadley Wickham (2011). Dates and Times Made Easy with lubridate.
  Journal of Statistical Software, 40(3), 1-25. URL https://www.jstatsoft.org/v40/i03/.

Hadley Wickham, Thomas Lin Pedersen and Dana Seidel (2023). scales: Scale Functions for
  Visualization. R package version 1.3.0. https://CRAN.R-project.org/package=scales